% Problems (Exercises) for Chapter 13

\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\subsection*{Easy}

\begin{problem}[PPCA vs PCA]
Contrast PCA and probabilistic PCA in assumptions and outputs.

\textbf{Hint:} Deterministic vs. probabilistic view; noise model; likelihood.
\end{problem}

\begin{problem}[Dimensionality Choice]
List two heuristics to select latent dimensionality.

\textbf{Hint:} Explained variance, information criteria.
\end{problem}

\begin{problem}[Gaussian Conditionals]
Recall the conditional of a joint Gaussian and its role in E-steps.

\textbf{Hint:} Use block-partitioned mean and covariance formulas.
\end{problem}

\begin{problem}[Rotation Ambiguity]
Explain why factor loadings are identifiable only up to rotation.

\textbf{Hint:} Orthogonal transforms preserve latent covariance.
\end{problem}

\begin{problem}[Missing Data]
Briefly describe how EM handles missing entries in linear-Gaussian models.

\textbf{Hint:} Expectation under observed mask.
\end{problem}

\begin{problem}[Whitening]
Define whitening and its relationship to PCA.

\textbf{Hint:} Diagonalize covariance and rescale.
\end{problem}

\subsection*{Medium}

\begin{problem}[PPCA Likelihood]
Derive the log-likelihood of PPCA and the M-step for $\sigma^2$.

\textbf{Hint:} Marginalize latents; differentiate w.r.t. variance.
\end{problem}

\begin{problem}[EM for FA]
Write the E and M steps for Factor Analysis with diagonal noise.

\textbf{Hint:} Use expected sufficient statistics of latents.
\end{problem}

\begin{problem}[Model Selection]
Compare AIC and BIC for choosing number of factors. Which tends to be more conservative?

\textbf{Hint:} Penalty strength differs.
\end{problem}

\begin{problem}[Sparse Loadings]
Propose a modification to encourage sparse factor loadings.

\textbf{Hint:} L1 penalties or spike-and-slab priors.
\end{problem}

\begin{problem}[Identifiability Constraints]
Suggest constraints to fix rotation/scaling without harming fit.

\textbf{Hint:} Upper-triangular loadings; orthonormal columns.
\end{problem}

\subsection*{Hard}

\begin{problem}[Equivalence of PPCA Solution]
Show that the MLE loading matrix spans the top-$k$ eigenvectors of the sample covariance.

\textbf{Hint:} Use spectral decomposition of covariance.
\end{problem}

\begin{problem}[FA vs. PPCA]
Demonstrate conditions where FA reduces to PPCA.

\textbf{Hint:} Isotropic noise and constrained loadings.
\end{problem}

\begin{problem}[Missing-not-at-random]
Discuss pitfalls when data are MNAR and how to adapt EM.

\textbf{Hint:} Model the missingness mechanism.
\end{problem}

\begin{problem}[Nonlinear Extension]
Sketch how to generalize linear factor models to VAEs.

\textbf{Hint:} Replace linear-Gaussian with neural encoder/decoder.
\end{problem}

\begin{problem}[Online EM]
Derive an online EM scheme for streaming data.

\textbf{Hint:} Maintain exponential moving averages of sufficient statistics.
\end{problem}


