% Chapter 10, Section 1

\section{Recurrent Neural Networks \difficultyInline{intermediate}}
\label{sec:rnns}

\subsection*{Intuition}

An RNN carries a running summary of the past—like a notepad you update after reading each word. This hidden state lets the model use prior context to influence current predictions. However, keeping reliable notes over long spans is hard: small errors can compound, and gradients may shrink or grow too much \cite{GoodfellowEtAl2016}.

\subsection*{Historical Context}

Early sequence models struggled with long-term dependencies due to vanishing gradients, motivating gated designs such as LSTM \cite{Hochreiter1997}. Practical training stabilized with techniques like gradient clipping and better initialization \cite{GoodfellowEtAl2016}.

% Index and glossary
\index{recurrent neural network}
\glsadd{recurrent-neural-network}

\subsection{Motivation}

Sequential data exhibits \emph{temporal dependencies} and \emph{order-sensitive} structure that cannot be modeled well by i.i.d. assumptions or fixed-size context windows alone \cite{GoodfellowEtAl2016,Prince2023,Bishop2006}. Examples include:
\begin{itemize}
    \item Time series: forecasting energy load, financial returns, or physiological signals (ECG).
    \item Natural language: the meaning of a word depends on its context; sentences conform to syntax and discourse structure.
    \item Speech/audio: phonemes combine to form words; coarticulation effects span multiple frames.
    \item Video: actions unfold over time; temporal cues disambiguate similar frames.
    \item Control and reinforcement learning: actions influence future observations, requiring memory.
\end{itemize}

Classic feedforward networks assume fixed-size inputs and lack a persistent state, making them ill-suited for long-range dependencies. Recurrent architectures introduce a hidden state that acts as a compact, learned memory \index{memory!neural} and enables conditioning on arbitrary-length histories. Historically, recurrent ideas trace back to early neural sequence models and dynamical systems; practical training matured with BPTT \cite{Rumelhart1986} and later with gated units to mitigate vanishing/exploding gradients \cite{Hochreiter1997}. For further background see the RNN overview in \cite{GoodfellowEtAl2016} and educational treatments in \cite{D2LChapterRNN,WebRNNWikipedia,WebDLBRNN}.

\index{sequence modeling}\index{temporal dependency}\glsadd{recurrent-neural-network}

\subsection{Basic RNN Architecture}

An RNN maintains a hidden state $\vect{h}_t$ that evolves over time:

\begin{align}
\vect{h}_t &= \sigma(\mat{W}_{hh} \vect{h}_{t-1} + \mat{W}_{xh} \vect{x}_t + \vect{b}_h) \\
\vect{y}_t &= \mat{W}_{hy} \vect{h}_t + \vect{b}_y
\end{align}

where $\vect{x}_t$ is input at time $t$, and $\sigma$ is typically tanh.

\paragraph{Visual aid.} The following unrolled diagram shows shared parameters across time:
\begin{figure}[h]
    \centering
    \begin{tikzpicture}[>=stealth, node distance=2.4cm]
        \tikzstyle{node}=[draw, rounded corners, minimum width=1.8cm, minimum height=1.1cm]
        \node[node] (h0) {$\vect{h}_{t-1}$};
        \node[node, right of=h0] (h1) {$\vect{h}_{t}$};
        \node[node, right of=h1] (h2) {$\vect{h}_{t+1}$};
        \draw[->] (h0) -- (h1) node[midway, above] {$\mat{W}_{hh}$};
        \draw[->] (h1) -- (h2) node[midway, above] {$\mat{W}_{hh}$};
        \node[below of=h0, yshift=1.2cm] (x0) {$\vect{x}_{t-1}$};
        \node[below of=h1, yshift=1.2cm] (x1) {$\vect{x}_{t}$};
        \node[below of=h2, yshift=1.2cm] (x2) {$\vect{x}_{t+1}$};
        \draw[->] (x0) -- (h0);
        \draw[->] (x1) -- (h1);
        \draw[->] (x2) -- (h2);
    \end{tikzpicture}
    \caption{Unrolled RNN with shared parameters across time steps.}
\end{figure}

\subsection{Unfolding in Time}

RNNs can be "unrolled" into a deep feedforward computation graph over time with \emph{shared parameters}. This perspective clarifies how gradients flow backward through temporal connections and why depth-in-time can cause vanishing/exploding gradients \cite{GoodfellowEtAl2016}.

\begin{equation}
\vect{h}_t = f(\vect{h}_{t-1}, \vect{x}_t; \vect{\theta})
\end{equation}

\paragraph{Visual aid.} Unfolding reveals repeated applications of the same transition function across steps. We annotate inputs, hidden states, and outputs to emphasize sharing and the temporal chain rule during BPTT.
\begin{figure}[h]
    \centering
    \begin{tikzpicture}[>=stealth, node distance=2.4cm]
        \tikzstyle{node}=[draw, rounded corners, minimum width=1.8cm, minimum height=1.1cm]
        \node[node] (h0) {$\vect{h}_{t-1}$};
        \node[node, right of=h0] (h1) {$\vect{h}_{t}$};
        \node[node, right of=h1] (h2) {$\vect{h}_{t+1}$};
        \draw[->] (h0) -- (h1) node[midway, above] {$f_{\vect{\theta}}$};
        \draw[->] (h1) -- (h2) node[midway, above] {$f_{\vect{\theta}}$};
        \node[below of=h0, yshift=1.2cm] (x0) {$\vect{x}_{t-1}$};
        \node[below of=h1, yshift=1.2cm] (x1) {$\vect{x}_{t}$};
        \node[below of=h2, yshift=1.2cm] (x2) {$\vect{x}_{t+1}$};
        \draw[->] (x0) -- (h0);
        \draw[->] (x1) -- (h1);
        \draw[->] (x2) -- (h2);
        \node[above of=h0, yshift=-1.2cm] (y0) {$\vect{y}_{t-1}$};
        \node[above of=h1, yshift=-1.2cm] (y1) {$\vect{y}_{t}$};
        \node[above of=h2, yshift=-1.2cm] (y2) {$\vect{y}_{t+1}$};
        \draw[->] (h0) -- (y0);
        \draw[->] (h1) -- (y1);
        \draw[->] (h2) -- (y2);
    \end{tikzpicture}
    \caption{Unrolling an RNN across time: the same parameters $\vect{\theta}$ are reused at each step.}
\end{figure}

This view connects RNNs to dynamic Bayesian networks and emphasizes that training complexity scales with the unroll length. See \cite{WebRNNWikipedia,GoodfellowEtAl2016,D2LChapterRNN}.

\index{unrolling in time}\index{shared parameters}

\subsection{Types of Sequences}

\textbf{One-to-one:} Fixed-size input to fixed-size output with temporal structure ignored or not present (e.g., static image classification with no time component).

\textbf{One-to-many:} Single input to sequence output (e.g., image captioning, where an image produces a sequence of words).

\textbf{Many-to-one:} Sequence input to single output (e.g., sentiment classification from a review; keyword spotting from audio).

\textbf{Many-to-many (synchronous):} Sequence labeling with aligned input/output lengths (e.g., part-of-speech tagging, CTC-free frame labeling).

\textbf{Many-to-many (asynchronous):} Sequence transduction with potentially different lengths (e.g., machine translation, speech recognition with CTC/attention). Attention helps bridge length mismatch by learning soft alignments \cite{Bahdanau2014}.

Design choices (teacher forcing, bidirectionality, attention, beam search) depend on whether future context is available and whether output timing must be causal. See \cite{D2LChapterRNN,WebRNNWikipedia} for further taxonomy.

\index{sequence types}\index{sequence labeling}\index{sequence transduction}

% Citations
See \cite{GoodfellowEtAl2016,Prince2023,Bishop2006,WebRNNWikipedia,WebDLBRNN,D2LChapterRNN} for introductions to sequence modeling and RNNs.

