% Chapter 9, Section 5

\section{Core CNN Algorithms \difficultyInline{intermediate}}
\label{sec:cnn-algorithms}

We introduce algorithms progressively, starting from basic cross-correlation to residual learning.

\subsection{Cross-Correlation and Convolution}
Given input $I\in\mathbb{R}^{H\times W\times C_{\text{in}}}$ and kernel $K\in\mathbb{R}^{k\times k\times C_{\text{in}}\times C_{\text{out}}}$, the output feature map $S\in\mathbb{R}^{H'\times W'\times C_{\text{out}}}$ under stride $s$ and padding $p$ is computed by cross-correlation as in \cref{sec:convolution}. Libraries often refer to this as "convolution" \cite{GoodfellowEtAl2016}.

\subsection{Backpropagation Through Convolution}
For loss $\mathcal{L}$ and pre-activation output $S = I * K$, the gradients are:
\begin{align}
\frac{\partial \mathcal{L}}{\partial K} &= I \star \frac{\partial \mathcal{L}}{\partial S},\\
\frac{\partial \mathcal{L}}{\partial I} &= \frac{\partial \mathcal{L}}{\partial S} * K^\text{rot},
\end{align}
where $\star$ denotes cross-correlation, $*$ denotes convolution, and $K^\text{rot}$ is the kernel rotated by $180^{\circ}$. Implementations use efficient im2col/FFT variants \cite{GoodfellowEtAl2016}.

\paragraph{Example (shape-aware):} For $I\in\mathbb{R}^{32\times32\times 64}$ and $K\in\mathbb{R}^{3\times3\times64\times128}$ with stride 1 and same padding, $S\in\mathbb{R}^{32\times32\times128}$. The gradient w.r.t. $K$ accumulates over spatial locations and batch.

\subsection{Pooling Backpropagation}
For max pooling, the upstream gradient is routed to the maximal input in each region; for average pooling, it is evenly divided among elements.

\subsection{Residual Connections}
In a residual block with input $\vect{x}$ and residual mapping $\mathcal{F}$, the output is $\vect{y}=\mathcal{F}(\vect{x})+\vect{x}$. Backpropagation yields $\frac{\partial \mathcal{L}}{\partial \vect{x}}=\frac{\partial \mathcal{L}}{\partial \vect{y}}\left(\frac{\partial \mathcal{F}}{\partial \vect{x}}+\mat{I}\right)$, stabilizing gradients and enabling very deep nets \cite{He2016}.

\subsection{Progressive Complexity: Depthwise Separable Convolutions}
Depthwise separable convolution factors standard convolution into depthwise (per-channel) and pointwise ($1\times1$) operations, reducing FLOPs and parameters (used by MobileNet). This keeps representational power while improving efficiency.

\paragraph{Parameter comparison.} For $C_{\text{in}}=C_{\text{out}}=c$ and kernel $k\times k$:
\begin{align}
\text{standard} &= k^2 c^2,\\
\text{depthwise separable} &= k^2 c + c^2,\quad \text{saving} \approx 1 - \frac{k^2 c + c^2}{k^2 c^2}.
\end{align}

\subsection{Normalization and Activation}
Batch normalization \cite{Ioffe2015} and ReLU-family activations improve optimization and generalization by smoothing the loss landscape and mitigating covariate shift.

\subsection{Other Useful Variants}
\begin{itemize}
    \item \textbf{Group/Layer Normalization:} alternatives when batch sizes are small.
    \item \textbf{Dilated Convolutions:} expand receptive field without pooling; effective in segmentation.
    \item \textbf{Anti-aliased Downsampling:} blur pooling or low-pass before stride to reduce aliasing artifacts in features.
\end{itemize}

\subsection*{Key Takeaways}
\begin{itemize}
    \item Cross-correlation is the practical "convolution" in deep learning; parameter sharing and locality are core.
    \item Residual connections preserve gradient flow, enabling very deep networks.
    \item Efficient convolutions (depthwise separable) trade compute for accuracy with strong Pareto gains.
    \item Normalization and careful downsampling are critical for stable optimization and preserving information.
\end{itemize}


