% Problems (Exercises) for Chapter 14

\section*{Problems}
\addcontentsline{toc}{section}{Problems}

\subsection*{Easy}
\begin{problem}[Undercomplete AE]
Explain why undercomplete AEs can avoid trivial identity mapping.
\textbf{Hint:} Bottleneck limits capacity.
\end{problem}

\begin{problem}[Denoising Noise]
How does noise type affect learned features?
\textbf{Hint:} Gaussian vs. masking vs. salt-and-pepper.
\end{problem}

\begin{problem}[Sparse Codes]
List benefits of sparsity in latent codes.
\textbf{Hint:} Interpretability, robustness, compression.
\end{problem}

\begin{problem}[Contractive Penalty]
What does a Jacobian penalty encourage?
\textbf{Hint:} Local invariance.
\end{problem}

\begin{problem}[Tied Weights]
Why tie encoder/decoder weights?
\textbf{Hint:} Parameter sharing, symmetry.
\end{problem}

\begin{problem}[Overfitting]
Name two signs your AE overfits.
\textbf{Hint:} Gap in reconstruction; poor transfer.
\end{problem}

\subsection*{Medium}
\begin{problem}[Loss Choices]
Compare MSE vs. cross-entropy for images.
\textbf{Hint:} Data scale/likelihood assumptions.
\end{problem}

\begin{problem}[Regularization Trade-offs]
Contrast denoising vs. contractive penalties.
\textbf{Hint:} Noise robustness vs. local smoothing.
\end{problem}

\begin{problem}[Linear AE]
Show linear AE with MSE learns PCA subspace.
\textbf{Hint:} Equivalence under constraints.
\end{problem}

\begin{problem}[Code Dimensionality]
Propose a method to choose latent dimension.
\textbf{Hint:} Validation on downstream task.
\end{problem}

\begin{problem}[Metric Learning]
Relate AE objectives to contrastive learning.
\textbf{Hint:} Geometry of embeddings.
\end{problem}

\subsection*{Hard}
\begin{problem}[Jacobian Penalty]
Derive gradient of contractive loss w.r.t. encoder parameters.
\textbf{Hint:} Chain rule through Jacobian norm.
\end{problem}

\begin{problem}[Denoising Bound]
Connect denoising training to score matching.
\textbf{Hint:} Small-noise limit.
\end{problem}

\begin{problem}[Regularization Failure]
Construct a case where regularization harms downstream performance.
\textbf{Hint:} Remove task-relevant fine detail.
\end{problem}

\begin{problem}[Generative Link]
Explain links between AEs and VAEs/flows.
\textbf{Hint:} Likelihood vs. reconstruction objectives.
\end{problem}

\begin{problem}[Visualization]
Design a LaTeX figure showing encoder–decoder with bottleneck.
\textbf{Hint:} Use tikz boxes and arrows.
\end{problem}


