% Chapter 4, Section 1: Overflow and Underflow

\section{Overflow and Underflow \difficultyInline{beginner}}
\label{sec:overflow-underflow}

\subsection{Intuition: The Problem with Finite Precision}

Imagine you're trying to measure the height of a building using a ruler that only has markings every meter. If the building is 10.7 meters tall, you might record it as 11 meters - you've lost some precision. Now imagine doing millions of calculations with this imprecise ruler, and you can see how small errors can compound into big problems.

Computers face a similar challenge. They can only represent a finite number of digits, so they must round numbers. This \textbf{finite precision} is like having a ruler with limited markings - some information is always lost.

In deep learning, this becomes critical because:
\begin{itemize}
    \item \textbf{Neural networks perform millions of calculations} - small errors accumulate
    \item \textbf{Exponential functions are common} - they can produce extremely large or small numbers
    \item \textbf{Gradients can become very small} - they might round to zero, breaking training
\end{itemize}

Computers represent real numbers with finite precision, typically using floating-point arithmetic. This leads to \textbf{rounding errors} that can accumulate and cause problems.

\subsection{Floating-Point Representation}

The IEEE 754 standard defines floating-point numbers. For 32-bit floats:
\begin{itemize}
    \item Smallest positive number: approximately $10^{-38}$
    \item Largest number: approximately $10^{38}$
    \item Machine epsilon: approximately $10^{-7}$
\end{itemize}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ylabel={Representable Numbers},
    xlabel={Value (log scale)},
    xmode=log,
    ymin=0,
    ymax=1,
    width=12cm,
    height=6cm,
    xtick={1e-40, 1e-30, 1e-20, 1e-10, 1, 1e10, 1e20, 1e30, 1e40},
    xticklabels={$10^{-40}$, $10^{-30}$, $10^{-20}$, $10^{-10}$, $1$, $10^{10}$, $10^{20}$, $10^{30}$, $10^{40}$},
    ytick=\empty
]
% Underflow region
\addplot[bookred, thick, domain=1e-50:1e-38] {0.2};
\node[bookred] at (axis cs: 1e-45, 0.3) {Underflow};

% Normal range
\addplot[bookpurple, thick, domain=1e-38:1e38] {0.5};
\node[bookpurple] at (axis cs: 1, 0.6) {Normal Range};

% Overflow region
\addplot[bookred, thick, domain=1e38:1e50] {0.8};
\node[bookred] at (axis cs: 1e45, 0.9) {Overflow};
\end{axis}
\end{tikzpicture}
\caption{Representable range for 32-bit floating-point numbers}
\label{fig:float-range}
\end{figure}

\subsection{Underflow}

\subsubsection{Intuition: When Numbers Become Too Small}

Think of underflow like trying to measure the width of a human hair with a ruler marked in meters. The hair is so thin that your ruler shows 0 meters - you've lost all information about the actual size.

In computers, \textbf{underflow} occurs when numbers become so small that they round to zero. This is like your ruler being too coarse to measure tiny objects.

\textbf{Underflow} occurs when numbers near zero are rounded to zero. This can be problematic when we need to compute ratios or logarithms. For example, the softmax function:

\begin{equation}
\text{softmax}(\vect{x})_i = \frac{\exp(x_i)}{\sum_j \exp(x_j)}
\end{equation}

can underflow if all $x_i$ are very negative.

\subsection{Overflow}

\subsubsection{Intuition: When Numbers Become Too Large}

Imagine trying to count the number of atoms in the universe using a calculator that can only display 8 digits. When you reach 99,999,999, the next number would be 100,000,000, but your calculator shows "Error" or resets to 0.

\textbf{Overflow} occurs when large numbers exceed representable values. In the softmax example, overflow can occur if some $x_i$ are very large.

\subsection{Numerical Stability}

To stabilize softmax, we use the identity:

\begin{equation}
\text{softmax}(\vect{x}) = \text{softmax}(\vect{x} - c)
\end{equation}

where $c = \max_i x_i$. This prevents both overflow and underflow.

Similarly, when computing $\log(\sum_i \exp(x_i))$, we use the \textbf{log-sum-exp} trick:

\begin{equation}
\log\left(\sum_i \exp(x_i)\right) = c + \log\left(\sum_i \exp(x_i - c)\right)
\end{equation}

\subsubsection{Example: Softmax Numerical Issues}

Consider computing softmax for $\vect{x} = [1000, 1001, 1002]$:

\textbf{Naive approach:}
\begin{align}
\exp(1000) &\approx \infty \quad \text{(overflow!)} \\
\exp(1001) &\approx \infty \quad \text{(overflow!)} \\
\exp(1002) &\approx \infty \quad \text{(overflow!)} \\
\text{softmax}(\vect{x}) &= [\text{NaN}, \text{NaN}, \text{NaN}]
\end{align}

\textbf{Stable approach:}
\begin{align}
c &= \max(1000, 1001, 1002) = 1002 \\
\exp(1000-1002) &= \exp(-2) \approx 0.135 \\
\exp(1001-1002) &= \exp(-1) \approx 0.368 \\
\exp(1002-1002) &= \exp(0) = 1.000 \\
\text{softmax}(\vect{x}) &= [0.090, 0.245, 0.665]
\end{align}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ylabel={Probability},
    xlabel={Class},
    ybar,
    ymin=0,
    ymax=0.7,
    width=10cm,
    height=6cm,
    xtick={1,2,3},
    xticklabels={Class 1, Class 2, Class 3}
]
\addplot coordinates {(1,0.090) (2,0.245) (3,0.665)};
\end{axis}
\end{tikzpicture}
\caption{Stable softmax computation for large inputs}
\label{fig:stable-softmax}
\end{figure}

\subsection{Other Numerical Issues}

\textbf{Catastrophic cancellation:} Loss of precision when subtracting nearly equal numbers.

\textbf{Accumulated rounding errors:} Small errors compound through many operations.

\textbf{Solutions:}
\begin{itemize}
    \item Use higher precision (64-bit floats)
    \item Algorithmic modifications (like log-sum-exp)
    \item Batch normalization
    \item Gradient clipping
\end{itemize}
