% Chapter 7, Section 3

\section{Early Stopping \difficultyInline{intermediate}}
\label{sec:early-stopping}

\textbf{Early stopping} monitors validation performance and stops training when it begins to degrade.

\subsection{Intuition: Stop Before You Memorize}

Imagine studying for an exam. Initially, practice improves your understanding (training and validation improve). If you keep cramming the exact same questions, you start memorizing answers that don't help with new questions (training improves, validation worsens). Early stopping is the principle of stopping at the point of best validation performance to avoid memorization.

\subsection{Algorithm}

\begin{enumerate}
    \item Train model and evaluate on validation set periodically
    \item Track best validation performance
    \item If no improvement for $p$ epochs (patience), stop
    \item Return model with best validation performance
\end{enumerate}

\begin{algorithm}[Early stopping meta-algorithm]
Let $n$ be the number of steps between evaluations.\\
Let $p$ be the patience (number of worsened validations before stopping).\\
Let $\vect{\theta}_0$ be the initial parameters.

\begin{enumerate}[leftmargin=*]
    \item $\vect{\theta} \leftarrow \vect{\theta}_0$, $i \leftarrow 0$, $j \leftarrow 0$, $v \leftarrow \infty$
    \item $\vect{\theta}^* \leftarrow \vect{\theta}$, $i^* \leftarrow i$
    \item \textbf{while} $j < p$ \textbf{do}
    \begin{enumerate}
        \item Update $\vect{\theta}$ by running the training algorithm for $n$ steps
        \item $i \leftarrow i + n$
        \item $v' \leftarrow \text{ValidationSetError}(\vect{\theta})$
        \item \textbf{if} $v' < v$ \textbf{then}
        \begin{enumerate}
            \item $j \leftarrow 0$
            \item $\vect{\theta}^* \leftarrow \vect{\theta}$, $i^* \leftarrow i$, $v \leftarrow v'$
        \end{enumerate}
        \item \textbf{else} $j \leftarrow j + 1$
        \item \textbf{end if}
    \end{enumerate}
    \item \textbf{end while}
    \item Return best parameters $\vect{\theta}^*$ and best step $i^*$
\end{enumerate}
\end{algorithm}

\subsection{Benefits}

\begin{itemize}
    \item \textbf{Simple and effective:} Requires only tracking validation performance and a patience parameter; widely used in practice \cite{GoodfellowEtAl2016,Prince2023}.
    \item \textbf{Automatically determines training duration:} Finds a good stopping time without a pre-fixed epoch budget, often saving substantial compute.
    \item \textbf{Implicit regularization:} Halting before convergence limits effective capacity by keeping weights smaller and preventing memorization; in some regimes it mimics an L2 constraint under gradient descent \cite{GoodfellowEtAl2016}.
    \item \textbf{Compatible with many settings:} Works with any loss, architecture (MLPs, CNNs, Transformers), and optimizer.
    \item \textbf{Reduces computational cost:} Training stops as soon as overfitting begins, reducing energy/time.
    \item \textbf{Improves generalization stability:} Curbs validation variance late in training when overfitting spikes.
\end{itemize}

\subsection{Considerations}

\begin{itemize}
    \item \textbf{Validation protocol:} Requires a reliable validation set and evaluation cadence; noisy metrics may trigger premature stops. Use smoothing or require monotone improvements.
    \item \textbf{Patience and frequency:} Patience $p$ and evaluation interval $n$ interact with LR schedules; too small $p$ can stop before a scheduled LR drop helps.
    \item \textbf{Checkpointing:} Always restore the best model (not the last); keep track of the weights at the best validation step.
    \item \textbf{Warmup and plateaus:} With warmup or long plateaus, consider larger patience or metric smoothing.
    \item \textbf{Multi-metric objectives:} For tasks with multiple metrics (e.g., accuracy and calibration), pick the primary metric or a composite.
    \item \textbf{Distributed training:} Ensure validation statistics are aggregated consistently across devices to avoid spurious decisions.
    \item \textbf{Historical context:} Early stopping predates modern deep learning and was popular in classical neural nets and boosting as a strong regularizer; it remains a standard baseline \cite{GoodfellowEtAl2016,Bishop2006}.
\end{itemize}

\begin{example}
\textbf{Example (vision):} Train a ResNet on CIFAR-10 with validation accuracy checked each epoch; use patience $p=20$. Accuracy peaks at epoch 142; training halts at 162 without improvement, and the checkpoint from 142 is used for testing.
\end{example}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.85]
    % axes
    \draw[->] (0,0) -- (9,0) node[right] {Epochs};
    \draw[->] (0,0) -- (0,4.5) node[above] {Loss};
    % training loss curve
    \draw[draw=bookred, thick] plot[smooth] coordinates {(0.3,4) (1,3.2) (2,2.6) (3,2.2) (4,1.9) (5,1.7) (6,1.5) (7,1.4) (8,1.35)};
    % validation loss curve (U-shaped)
    \draw[draw=bookpurple, thick] plot[smooth] coordinates {(0.3,3.9) (1,3.1) (2,2.5) (3,2.15) (4,1.95) (5,1.92) (6,1.95) (7,2.05) (8,2.2)};
    % vertical line at best epoch
    \draw[dashed] (4,0) -- (4,4.5) node[above] {Best epoch};
    \node[text=bookpurple] at (6.9,2.3) {Validation};
    \node[text=bookred] at (7.3,1.35) {Training};
\end{tikzpicture}
\caption{Early stopping: validation loss reaches a minimum before training loss; the best checkpoint is saved and restored.}
\label{fig:early-stopping-curve}
\end{figure}

% Index entries
\index{early stopping}
\index{validation set}

