% Chapter 6: Deep Feedforward Networks

\chapter{Deep Feedforward Networks}
\label{chap:feedforward-networks}

This chapter introduces deep feedforward neural networks, also known as multilayer perceptrons (MLPs). These are the fundamental building blocks of deep learning.

\section*{Learning Objectives}
\label{sec:learning-objectives}

After studying this chapter, you will be able to:

\begin{enumerate}
    \item \textbf{Understand the architecture} of feedforward neural networks and how they process information from input to output.
    
    \item \textbf{Master activation functions} and their role in introducing non-linearity, including the trade-offs between different choices.
    
    \item \textbf{Design appropriate output layers} and loss functions for different types of machine learning tasks (regression, binary classification, multiclass classification).
    
    \item \textbf{Derive and implement backpropagation} to efficiently compute gradients for training neural networks.
    
    \item \textbf{Make informed design choices} about network architecture, initialization, and training procedures.
    
    \item \textbf{Apply theoretical knowledge} to solve practical problems involving feedforward networks.
\end{enumerate}




\input{chapters/chap06-sec01}
\input{chapters/chap06-sec02}
\input{chapters/chap06-sec03}
\input{chapters/chap06-sec04}
\input{chapters/chap06-sec05}
\input{chapters/chap06-real-world-applications}
\input{chapters/chap06-sec06}

% Chapter summary and problems
\input{chapters/chap06-key-takeaways}
\input{chapters/chap06-problems}
